---
title: "cog_2yr"
author: "Julia Gallucci"
date: "20/08/2025"
output: html_document
---
Load necessary packages
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidyverse)
library(afex)       # For repeated measures ANOVA
library(emmeans)    # For post-hoc comparisons

```

Load cog longitudinal data
```{r}
#X2yr_cognition is data from 2-year follow up, combine with baseline data
#all combines discovery and replication to maximize sample
cog <- left_join(all,X2yr_cognition)

#select relevant columns
cog <- cog %>%
  dplyr::select(src_subject_id,
                groups, 
                nihtbx_flanker_uncorrected,
                nihtbx_flanker_uncorrected_2yr,
                nihtbx_list_uncorrected,
                nihtbx_list_uncorrected_2yr,
                nihtbx_cardsort_uncorrected,
                nihtbx_cardsort_uncorrected_2yr,
                nihtbx_pattern_uncorrected,
                nihtbx_pattern_uncorrected_2yr,
                nihtbx_picture_uncorrected,
                nihtbx_picture_uncorrected_2yr,
                nihtbx_reading_uncorrected,
                nihtbx_reading_uncorrected_2yr,
                nihtbx_picvocab_uncorrected,
                nihtbx_picvocab_uncorrected_2yr,
                pea_wiscv_trs,
                pea_wiscv_trs_2yr,
                rey_all_correct,
                rey_all_correct_2yr,
                lmt_scr_num_correct,
                lmt_scr_num_correct_2yr)

```

Pivot to long format
```{r}
baseline_cols <- names(cog)[!(names(cog) %in% c("src_subject_id", "groups")) & !grepl("_2yr$|_4yr$", names(cog))]
year2_cols <- paste0(baseline_cols, "_2yr")
#year4_cols <- paste0(baseline_cols, "_4yr")

cog_long <- cog %>%
  pivot_longer(
    cols = all_of(c(baseline_cols, year2_cols)),
    names_to = "measure_time",
    values_to = "score"
  ) %>%
  mutate(
    time = case_when(
      grepl("_2yr$", measure_time) ~ "year2",
      TRUE ~ "baseline"
    ),
    measure = gsub("_2yr$", "", measure_time)
  ) %>%
  select(src_subject_id, groups, measure, time, score) %>%
  mutate(
    time = factor(time, levels = c("baseline", "year2")),
    groups = factor(groups)
  )
```

NonPLE reference group
```{r}
cog_nonPLE <- left_join(nonPLE, X2yr_cognition, by = "src_subject_id")
cog_nonPLE$groups <- "NonPLE"

cog_nonPLE <- cog_nonPLE %>%
  select(src_subject_id, groups, all_of(c(baseline_cols, year2_cols)))

cog_long_nonPLE <- cog_nonPLE %>%
  pivot_longer(
    cols = all_of(c(baseline_cols, year2_cols)),
    names_to = "measure_time",
    values_to = "score"
  ) %>%
  mutate(
    time = case_when(
      grepl("_2yr$", measure_time) ~ "year2",
      TRUE ~ "baseline"
    ),
    measure = gsub("_2yr$", "", measure_time)
  ) %>%
  select(src_subject_id, groups, measure, time, score) %>%
  mutate(
    time = factor(time, levels = c("baseline", "year2")),
    groups = factor(groups)
  )

```

Composite score

Calculates the mean and standard deviation for each cognitive measure at each time point in NonPLE. Then computes z-scores for all subjects relative to the NonPLE reference at the same time point. For each subject, averages the z-scores across all cognitive measures at a given time to create a single composite cognitive score.
```{r}
# Combine all data
cog_long_all <- bind_rows(cog_long, cog_long_nonPLE)

# Reference mean & SD for NonPLE at each time point
ref_stats_time <- cog_long_nonPLE %>%
  group_by(measure, time) %>%
  summarise(
    mean_ref = mean(score, na.rm = TRUE),
    sd_ref   = sd(score, na.rm = TRUE),
    .groups = "drop"
  )

# Compute z-scores relative to NonPLE at each time point
cog_long_scaled_ref <- cog_long_all %>%
  left_join(ref_stats_time, by = c("measure", "time")) %>%
  mutate(score_z_ref = (score - mean_ref) / sd_ref)

# Compute composite per subject/time
cog_composite_ref <- cog_long_scaled_ref %>%
  group_by(src_subject_id, groups, time) %>%
  summarise(composite_ref = mean(score_z_ref, na.rm = TRUE), .groups = "drop")

# Remove subjects with any missing composite scores
cog_composite_ref <- cog_composite_ref %>%
  group_by(src_subject_id) %>%
  filter(!any(is.na(composite_ref))) %>%
  ungroup()

# How many subjects per group

group_counts <- cog_composite_ref %>%
  distinct(src_subject_id, groups) %>%
  group_by(groups) %>%
  summarise(n = n())
group_counts

# Cluster 1 - 156
# Cluster 2 - 388
# Cluster 3 - 253
# Cluster 4 - 445
# Cluster 5 - 323
# NonPLE - 2117

```

Repeated measures ANOVA
```{r}
# Ensure factors
cog_composite_ref$time <- factor(cog_composite_ref$time, levels = c("baseline", "year2"))
cog_composite_ref$groups <- factor(cog_composite_ref$groups)

# Run ANOVA
aov_res <- aov_ez(
  id = "src_subject_id",
  dv = "composite_ref",
  data = cog_composite_ref,
  within = "time",
  between = "groups"
)

summary(aov_res) #significant group x time interaction

#get pairwise interactions
emm <- emmeans(aov_res, pairwise ~ groups | time)
# Pairwise comparisons with fdr adjustment
pairwise_contrasts <- contrast(emm, method = "pairwise", adjust = "fdr")
pairwise_contrasts

```

```{r}
# Group colors
default_colors <- scales::hue_pal()(5)
# Plot clusters with shaded confidence intervals, fixed y-axis, no NonPLE line
ggplot(cog_composite_ref %>% filter(groups != "NonPLE"), 
       aes(x = time, y = composite_ref, color = groups, group = groups, fill = groups)) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.2, color = NA) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  annotate("text", x = 2, y = 0.05, label = "NonPLE", hjust = -0.5, vjust = 0.5, color = "grey") +
  scale_color_manual(values = c(
    "1" = default_colors[1],
    "2" = default_colors[2],
    "3" = default_colors[3],
    "4" = default_colors[4],
    "5" = default_colors[5]
  )) +
  scale_fill_manual(values = c(
    "1" = default_colors[1],
    "2" = default_colors[2],
    "3" = default_colors[3],
    "4" = default_colors[4],
    "5" = default_colors[5]
  )) +
  coord_cartesian(ylim = c(-2, 2)) +
  theme_classic() +
  labs(
    y = "Composite Cognitive Performance Z-Score",
    x = "Time",
    color = "Cluster",
    fill = "Cluster"
  )

```
